### 2020-05-13 probabilisticProgrammingMartingales.tex
ongoing

When proving almost sure termination in this way /akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to. Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.

Non-probabilistis lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter. In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction. There are, however, some restricted variations on the reduction order that do not have this problem.

\paragraph{}
Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one. To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.

A position \akr{I'm not really sure how to phrase the citation, but this is adapded from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The subterm at a position is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $xy \mid \lambda ; \cdot$ is undefined). A position such that $M\mid P$ does exist is said to occur in $M$. Substitution at a given position, $M[N/P]$, is defined similarly.

@@

With this notation, a more general reduction relation can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}

\paragraph{}
Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x. \underline x 0 \underline + x \underline 0) (\lambda y. \tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$. It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. This is itself inadequate because some of the positions in different reachable terms need to be considered the same.

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence. Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation. The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
\begin{enumerate}
    \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N | P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp. second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp. $3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
    \item If $N \mid P = (\lambda x. A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)$
\end{enumerate}

@@

In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures. I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}.

>>>>

When proving almost sure termination in this way \akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to. Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.

Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter. In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction. There are, however, some restricted variations on the reduction order that do not have this problem.

\paragraph{}
Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one. To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.

A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The subterm at a position is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). A position such that $M\mid P$ does exist is said to occur in $M$. Substitution at a given position, $M[N/P]$, is defined similarly.

@@

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
\lo{On first reading, I struggled to parse the following clauses.}
\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a position $P$ that occurs in the term $M$ in question.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}
\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}

\paragraph{}
Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x. {\underline x \, 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y. \tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$. It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. This is itself inadequate because some of the positions in different reachable terms need to be considered the same.

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence. Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation. \lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and relations $\sim_p$ and $\sim_c$, here, before the formal definitions. Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
\begin{enumerate}
    \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp. second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp. $3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
    \item If $N \mid P = (\lambda x. A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
    \item If $N \mid P = (\tY \lambda x. A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, 
    \[
    (O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)
    \]
\end{enumerate}

@@

In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures. I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
\lo{I can suggest possible \LaTeX\ packages for drawing. Can you draw them and upload a scan?}
